Grove: A Bidirectionally Typed Collaborative Structure Editor Calculus

### 2024-11-22 sections/marking-system.tex
directory restructuring and hiding appendices

Resolving conflicts is often a time-consuming process 
and requires understanding the semantics of the program
even while it still has conflicts.

Traditionally, conflicts are indicated by inserting extra-linguistic conflict markers into files. 
These markers can limit the operation of language services that need a well-formed or well-typed term, 
e.g., type error reporting.

Ignoring the markers is not sufficient because 
multiple conflicting versions of the code may appear
between the markers. This can complicate variable resolution and type checking.

@@

We follow the marked lambda calculus in basing our \emph{marked grove calculus} in bidirectional type checking~\cite{dunfield2019} and deploying gradual typing~\cite{siek2015} when conflicts do not allow a single type to be inferred, then layer on a unification-based type inference system to opportunistically fill holes or suggest partial solutions when there are conflicting types due to conflicting syntax. We give an overview of the key judgments and rules in \autoref{sub:marking-groves} but due to space considerations and because much of the development is based directly on the prior work, we leave the full details to the supplemental material and the mechanized metatheory, which is briefly outlined in \ref{sub:marking-agda}.

@@

The first rule invokes the matched arrow judgment (which handles the situation where the expected type is unknown \cite{snapl15}) and performs a type consistency check with the argument type annotation, after which we can analyze the body against the expected output type while recursively marking it.

@@

\subsubsection{Syntax}
\label{sub:unmarked-lang}
\autoref{fig:marking-syntax} introduces the syntax of the marked grove calculus. We define an \emph{unmarked} language, which comprises unmarked expressions, $\EMV$,  child expressions, $\ELV$, types, $\TMV$, and  child types $\TLV$.
We also define a corresponding \emph{marked} language,  which comprises marked expressions, $\ECMV$, child expressions, $\ELMV$, types, $\TCMV$, and child types, $\TCLV$. This syntax is an instantiation of the generic term syntax introduced in \autoref{sub:Groves}.
The marked language differs from the unmarked language only by the inclusion of the marks that arise due to type errors taken directly from the prior work.

@@

Once marked, the corresponding marked expression typing judgments, $\synConstraint{\ctx}{\ECMV}{\PTMV}{\cConstraint}{\mConstraint}$, $\synConstraint{\ctx}{\ELMV}{\PTMV}{\cConstraint}{\mConstraint}$, 
$\anaConstraint{\ctx}{\ECMV}{\PTMV}{\cConstraint}{\mConstraint}$, and $\anaConstraint{\ctx}{\ELMV}{\PTMV}{\cConstraint}{\mConstraint}$ emit type inference constraints, $\cConstraint$, as in prior work, and, recursively collect location conflict contexts, $\mConstraint$, discussed further below. When the system generates inference constraints and tries to unify them, it encounters unknown types that need to be distinguished. This motivates the need for $\PTMV$ in \autoref{fig:marking-syntax}, which extends the sort of semantic types to enable linking generated unknown types to their associated provenance denoted by $q$. Provenances help locate the origin of the unknown type by associating them with their location $\ell$, edge-ID(w), or vertex-ID (u).

@@

    \inferrule[MALocalConflict]{ 
        \Setof{\anaConstraint{\ctx}{\ECMV_i}{\PTMV}{\cConstraint_i}{\mConstraint_i}}_{i<n}
    }{ 
      \anaConstraint{\ctx}{\conflict{\ECMV}}{\PTMV}{\bigcup_{i<n} \cConstraint_i \cup
      \Setof{\constrain{\TUnknown^{\mathsf{exp}(\ell)}}{\PTMV}}}{\bigcup_{i<n} \mConstraint_i}
    }
\end{mathpar}

>>>>

Resolving conflicts is often a time-consuming process 
and requires understanding the semantics of the program
even while it still has conflicts. Traditionally, conflicts are indicated by inserting extra-linguistic conflict markers into files.

These markers can limit the operation of language services that need a well-formed or well-typed term, 
e.g., type error reporting. Ignoring the markers is not sufficient because 
multiple conflicting versions of the code may appear between the markers. This can complicate variable resolution and type checking.

@@

We follow the marked lambda calculus in basing our \emph{marked grove calculus} in bidirectional type checking~\cite{10.1145/3450952} and deploying gradual typing~\cite{siek2015} when conflicts do not allow a single type to be inferred, then layer on a unification-based type inference system to opportunistically fill holes or suggest partial solutions when there are conflicting types due to conflicting syntax. We give an overview of the key judgments and rules in \autoref{sub:marking-groves} but due to space considerations and because much of the development is based directly on the prior work, we leave the full details to the supplemental material and the mechanized metatheory, which is briefly outlined in \ref{sub:marking-agda}.

@@

The first rule invokes the matched arrow judgment (which handles the situation where the expected type is unknown \cite{siek2015}) and performs a type consistency check with the argument type annotation, after which we can analyze the body against the expected output type while recursively marking it.

@@

\subsubsection{Syntax}
\label{sub:unmarked-lang}
\autoref{fig:marking-syntax} introduces the syntax of the marked grove calculus. We define an \emph{unmarked} language, which comprises unmarked expressions, $\EMV$,  child expressions, $\ELV$, types, $\TMV$, and  child types $\TLMV$.
We also define a corresponding \emph{marked} language,  which comprises marked expressions, $\ECMV$, child expressions, $\ELMV$, types, $\TCMV$, and child types, $\TCLV$. This syntax is an instantiation of the generic term syntax introduced in \autoref{sub:Groves}.
The marked language differs from the unmarked language only by the inclusion of the marks that arise due to type errors taken directly from the prior work.

@@

Once marked, the corresponding marked expression typing judgments, $\synConstraint{\ctx}{\ECMV}{\PTMV}{\cConstraint}{\mConstraint}$, $\synConstraint{\ctx}{\ELMV}{\PTMV}{\cConstraint}{\mConstraint}$, 
$\anaConstraint{\ctx}{\ECMV}{\PTMV}{\cConstraint}{\mConstraint}$, and $\anaConstraint{\ctx}{\ELMV}{\PTMV}{\cConstraint}{\mConstraint}$ emit type inference constraints, $\cConstraint$, as in prior work, and, recursively collect location conflict contexts, $\mConstraint$, discussed further below. When the system generates inference constraints and tries to unify them, it encounters unknown types that need to be distinguished. This motivates the need for $\PTMV$ in \autoref{fig:marking-syntax}, which extends the sort of semantic types to enable linking generated unknown types to their associated provenance denoted by $q$. Provenances help locate the origin of the unknown type by associating them with their location ($\ell$), edge-ID ($w$), or vertex-ID ($u$).

@@

    \inferrule[MALocalConflict]{ 
        \Setof{\anaConstraint{\ctx}{\ECMV_i}{\PTMV}{\cConstraint_i}{\mConstraint_i}}_{i<n}
    }{ 
      \anaConstraint{\ctx}{\conflict{\ECMV}}{\PTMV}{\bigcup_{i<n} \cConstraint_i}{\bigcup_{i<n} \mConstraint_i}
    }
\end{mathpar}

### 2024-11-22 sections/related-work.tex
directory restructuring and hiding appendices

\subsection{Patch Languages}
There have been many different designs for patch languages and indeed many imperative data structures and their associated operations can be construed as patch languages in the most general sense. In the context of collaborative coding, patch languages can differ in the data structures they operate on (e.g. line-based text~\cite{DBLP:conf/haskell/Roundy05}, character-based text~\cite{DBLP:conf/sigmod/EllisG89,DBLP:journals/pacmhci/LittLKH22}, tree-structured data). 
They can also differ in how they identify locations within the data (e.g. by using numeric offsets~\cite{DBLP:conf/sigmod/EllisG89}, unique identifiers, or paths through a tree).
Finally, patch languages differ in which specific actions are supported explicitly. Insertion and deletion are common, while code relocation, copying, undo, and other operations are variously also included.

@@

\subsection{Merging}
\subsubsection{Operational Transforms}
The most common approach to merging concurrently developed patches is to deploy an operational transform~\cite{DBLP:conf/sigmod/EllisG89} whereby locations in a remote patch are modified based on the action of a local patch. Standard three-way merge algorithms in version control systems deploy this approach, as do real-time collaborative editors. There have been a number of papers studying the algebraic properties of merging patches in this style.

For example, the Darcs~\cite{DBLP:conf/haskell/Roundy05} version control system, like Grove, represents repositories using sets of patches. Using operational transforms, Darcs achieves commutativity in many cases, but not between conflicting patches. The theory of Darcs defines and algebraically characterizes when operational transforms do satisfy the properties of associativity and commutativity. Recent work on homotopical patch theory \cite{DBLP:journals/jfp/AngiuliMLH16} has similarly developed an abstract algebraic framework for distinguishing sensible merges.

@@

Other examples of CRDT-based editors include and Peritext~\cite{DBLP:journals/pacmhci/LittLKH22} and Zed~\cite{zed-blog}. These systems differ from Grove by targeting synchronous editing rather than version control and by operating on sequential text rather than tree structures.

There have been other recent efforts to develop CmRDTs for tree data structures~\cite{DBLP:conf/icdcs/PreguicaMSL09, Highly-Available}. However, the focus in this work has been on avoiding conflicts and cycles entirely by applying \emph{ad hoc} heuristics for conflict resolution at merge-time, e.g. using reported timestamps or favoring particular directions in the tree. Our approach instead embraces manual conflict resolution, as is common practice in software projects where arbitrarily losing code is not acceptable.

>>>>

\subsection{Patch Languages}
There have been many different designs for patch languages and indeed many imperative data structures and their associated operations can be construed as patch languages in the most general sense. In the context of collaborative coding, patch languages can differ in the data structures they operate on (e.g. line-based text~\cite{DBLP:conf/haskell/Roundy05}, character-based text~\cite{DBLP:conf/sigmod/EllisG89,DBLP:journals/pacmhci/LittLKH22}, tree-structured data~\cite{DBLP:conf/esa/Klein98}). 
They can also differ in how they identify locations within the data (e.g. by using numeric offsets~\cite{DBLP:conf/sigmod/EllisG89}, unique identifiers, or paths through a tree).
Finally, patch languages differ in which specific actions are supported explicitly. Insertion and deletion are common, while code relocation, copying, undo, and other operations are variously also included.

@@

\subsection{Merging}
\subsubsection{Operational Transforms}
The most common approach to merging concurrently developed patches is to deploy an operational transform~\cite{DBLP:conf/sigmod/EllisG89} whereby locations in a remote patch are modified based on the action of a local patch. Standard three-way merge algorithms in version control systems deploy this approach, as do real-time collaborative rich text editors~\cite{DBLP:journals/concurrency/IgnatAO21}.

There have been a number of papers studying the algebraic properties of merging patches in this style. For example, the Darcs~\cite{DBLP:conf/haskell/Roundy05} version control system, like Grove, represents repositories using sets of patches. Using operational transforms, Darcs achieves commutativity in many cases, but not between conflicting patches. The theory of Darcs defines and algebraically characterizes when operational transforms do satisfy the properties of associativity and commutativity. Recent work on homotopical patch theory \cite{DBLP:journals/jfp/AngiuliMLH16} has similarly developed an abstract algebraic framework for distinguishing sensible merges.

@@

Much of the prior work on CRDT-based collaborative editing has focused on text editing 
\cite{DBLP:conf/icdcs/PreguicaMSL09, 
DBLP:conf/cscw/OsterUMI06,
DBLP:journals/jpdc/RohJKL11,
DBLP:conf/wikis/Grishchenko10,
DBLP:conf/icdcs/WeissUM09,
DBLP:conf/doceng/NedelecMMD13,
DBLP:conf/group/NicolaescuJDK16,ahmed2011evaluating,kleppmann2020moving}, typically with text tagged with unique IDs. These techniques have been used to develop collaborative text editors, e.g. Zed~\cite{zed-blog}.

Peritext develops a CRDT-based approach for collaboratively editing rich text, which is structured as text with hierarchically marked regions~\cite{DBLP:journals/pacmhci/LittLKH22}.

There have been other recent efforts to develop CmRDTs for richer tree data structures~\cite{DBLP:conf/icdcs/PreguicaMSL09, Highly-Available}. However, the focus in this work has been on avoiding conflicts and cycles entirely by applying \emph{ad hoc} heuristics for conflict resolution at merge-time, e.g. using reported timestamps or favoring particular directions in the tree. Our approach instead embraces manual conflict resolution, as is common practice in software projects where arbitrarily losing code is not acceptable.
